---
title: "NHANES database build-time checks"
author: "Deepayan Sarkar"
date: "`r Sys.Date()`"
output: github_document
editor_options: 
  chunk_output_type: console
---


## Tables missing from table metadata

The following tables are in the database but not in the metadata stored in the DB.


```{r}
## code to generate list
nhanes_url <- function (nh_table) {
  nh_year <- nhanesA:::.get_year_from_nh_table(nh_table)
  paste0(nhanesA:::nhanesURL, nh_year, "/", nh_table, ".htm")
}
library(nhanesA) # mainly for querying database
all_tables_in_db <- 
  subset(nhanesA:::.nhanesQuery("select * from information_schema.tables;"),
         TABLE_SCHEMA == "Translated")$TABLE_NAME
all_tables_in_metadata <- 
  nhanesA:::.nhanesQuery("select * from Metadata.QuestionnaireDescriptions")$TableName
URLs <- 
  (setdiff(all_tables_in_db, all_tables_in_metadata) 
   |> grep(pattern = "^P_", invert = TRUE, value = TRUE) 
   |> sapply(FUN = nhanes_url))
```

```{r results="asis", comment = ""}
cat("", sprintf("- [%s](%s)", names(URLs), URLs), sep = "\n")
```

This needs to be looked at manually (many, but not all, are due to obvious bugs 
in the NHANES documentation pages).


## Cross check tables with NHANES master list

The NHANES website has a master list of all tables at

<https://wwwn.cdc.gov/Nchs/Nhanes/search/DataPage.aspx>

Let's download this and save it in a CSV file using Python...


```{python, results='hide'}
from bs4 import BeautifulSoup
import requests

## extraction functions for td elements
def etext(obj): return obj.get_text().strip()
def eurl(obj): return obj.find('a').get_attribute_list('href')[0]

url = 'https://wwwn.cdc.gov/Nchs/Nhanes/search/DataPage.aspx'

source_html = requests.get(url).content.decode('utf-8')
soup = BeautifulSoup(source_html, 'html.parser')

_table = soup.find('table', {'id' : 'GridView1'}) 

f = open('table_manifest.csv', 'w')
f.write("Table,Years,PubDate,DocURL,DataURL\n")

for row in _table.tbody.find_all('tr'):
    [year, docfile, datafile, pubdate] = row.find_all('td')
    if etext(pubdate) != 'Withdrawn':
        f.write("%s,%s,%s,https://wwwn.cdc.gov%s,https://wwwn.cdc.gov%s\n" % 
                    (etext(docfile).split()[0], 
                     etext(year), 
                     etext(pubdate),
                     eurl(docfile),
                     eurl(datafile)))
f.close()
```

...and then read this file in using R.

```{r}
manifest <- read.csv("table_manifest.csv")
## fixup some weird URLs
manifest <- within(manifest, {
  DocURL <- gsub("cdc.gov../", "cdc.gov/Nchs/Nhanes/", DocURL, fixed = TRUE)
  DataURL <- gsub("cdc.gov../", "cdc.gov/Nchs/Nhanes/", DataURL, fixed = TRUE)
})
```

Sanity checks:

```{r}
with(manifest, table(tools::file_ext(DocURL)))   # doc file extensions
with(manifest, table(tools::file_ext(DataURL)))  # data file extensions
with(manifest, Table[duplicated(Table)])         # uniqueness of table names
```

The `aspx` files and duplicated table name comes from additional details 
specific to missingness in DXA and OMB tables (see links). Note that 
`nhanesA` handles "DXA" specially, e.g., `nhanesA::nhanesDXA()`, but not OMB.


```{r}
subset(manifest, Table == "All")
```

We will simply skip these two entries.

```{r}
manifest <- subset(manifest, Table != "All")
```


The ZIP extensions are from

```{r}
subset(manifest, tolower(tools::file_ext(DataURL)) == "zip")
```

These are very large files, with multiple entries per subject storing 
minute-by-minute results recorded in a physical activity monitoring device. 
These data are not included in the database (and probably should not be), but 
we will retain these rows to remind us that the tables exist.

OK, let's now see if any of these are missing either from the list of tables
in the database, or from the table metadata.


```{r}
setdiff(manifest$Table, all_tables_in_db) # missing from DB
```

Links for exploration:

```{r results='asis'}
paste0("- <", 
       subset(manifest, Table %in% setdiff(manifest$Table, all_tables_in_db))$DocURL,
       ">") |>
  cat(sep = "\n")
```

Tables missing from metadata:

```{r}
setdiff(manifest$Table, all_tables_in_metadata) # missing from metadata
```

Links for exploration (excluding `P_*` tables for now): [Note that these should
be a superset of the earlier list of tables missing from the metadata]

```{r results='asis'}
paste0("- <", 
       subset(manifest, !startsWith(Table, "P_") & 
                Table %in% setdiff(manifest$Table, all_tables_in_metadata))$DocURL,
       ">") |>
  cat(sep = "\n")
```

Although there should not be, let's also make sure that no table in the DB is 
missing from the manifest.

```{r}
setdiff(all_tables_in_db, manifest$Table)
```


## Missing documentation

For some reason, many tables in the current metadata do not have a URL. 

```{r}
table_metadata <- nhanesA:::.nhanesQuery("select * from Metadata.QuestionnaireDescriptions")
subset(table_metadata, !startsWith(DocFile, "https"))[-1]
```

For now, we will not try to conjecture why these have failed.

Let's check a few other things regarding the URLs:

```{r}
manifest_doc_url <- with(manifest, structure(DocURL, names = Table))
manifest_data_url <- with(manifest, structure(DataURL, names = Table))
metadata_doc_url <- with(subset(table_metadata, DocFile != ""), structure(DocFile, names = TableName))
metadata_data_url <- with(subset(table_metadata, DataFile != ""), structure(DataFile, names = TableName))
```

- If not missing, doc / data URLs in the metadata should match those in the manifest.

```{r}
all(metadata_doc_url == manifest_doc_url[names(metadata_doc_url)])
all(metadata_data_url == manifest_data_url[names(metadata_data_url)])
```

- Auto-generated Doc URLs should match the ones in the manifest

```{r}
generated_urls <- sapply(names(manifest_doc_url), nhanes_url)
subset(data.frame(table = names(manifest_doc_url), 
                  generated_urls, 
                  manifest_doc_url), 
       generated_urls != manifest_doc_url)
```

At the time of writing, `PFC_POOL` and `SSNH4THY` give mismatches because they 
are from cycle 2 but do not have the `_B` suffix, so the auto-generated URLs 
have the wrong year. These should probably be special-cased in `nhanesA`.

The others are legitimate differences. For these tables, the master table on 
the NHANES website links to a note, but the underlying URLs are still valid.


## Translation

Do we want to (optionally) translate special codes in numeric values?

```{r}
varcb <- nhanesA:::.nhanesQuery("select * from Metadata.VariableCodebook")
dim(varcb)
head(varcb)
```

First of all, is there a way to "know" that a variable is supposed to be 
numeric? 

```{r}
(num_variables <- with(varcb, length(unique(Variable)))) # total number of variables
with(varcb, xtabs(~ ValueDescription)) |> sort(decreasing = TRUE) |> head(20)
```

So hopefully the presence of `"Range of Values"` as an option for a variable 
should indicate that it is a continuous variable. Let's check for any typos just 
in case:

```{r}
with(varcb,
     grep("range", ValueDescription, ignore.case = TRUE, value = TRUE)
     |> table()
     |> sort(decreasing = TRUE))
```

(suggests other problems, but not in what we care about), and

```{r}
with(varcb,
     grep("values", ValueDescription, ignore.case = TRUE, value = TRUE)
     |> table()
     |> sort(decreasing = TRUE))
```

OK, so let's now see what other values appear for these 'numeric' variables.

```{r}
numeric_variables <- 
  unique(subset(varcb, 
                ValueDescription == "Range of Values")[["Variable"]])
length(numeric_variables)
numeric_descriptions <- 
  subset(varcb, 
         Variable %in% numeric_variables, 
         select = c(Variable, ValueDescription, CodeOrValue))
combinations <- (
  xtabs(~ ValueDescription + CodeOrValue, numeric_descriptions, 
        subset = ValueDescription != "Range of Values") 
  |> as.data.frame.table() 
  |> subset(Freq > 0)
)
dim(combinations)
options(width = 120)
print(combinations, max = 2000)
```

OK, so why is "English and Spanish" included here?

```{r}
which_vars <- 
  unique(subset(numeric_descriptions, 
                ValueDescription == "English and Spanish")$Variable)
subset(varcb, Variable %in% which_vars & 
         ValueDescription == "Range of Values")
```

These seem to be bugs in the NHANES data.

```{r}
nhanes("DS1IDS_G")$DR1LANG |> table()
```

For comparison,

```{r}
nhanes("DS1IDS_E")$DR1LANG |> table()
```

Can't see an easy way to automate this. 

But in general, for numeric variables, there seem to be mainly two types of 
records we need to worry about:

- Missing values with a specific reason for being missing

- Left or right censored values (less than or greater than something)

Also, 0 codes are somewhat tricky.






