---
title: "Variables that are both numeric and categorical"
output: github_document
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(comment = "",
                      fig.path = "figures/",
                      out.width = "100%",
                      fig.width = 12, fig.height = 8,
                      dev = "svglite", dev.args = list(pointsize = 12),
                      cache = TRUE,
                      cache.path = "./knitr-cache/numcat/")
options(width = 100)
library(lattice)
library(kableExtra) # for kable()
## library(DT)
```


Some variables are currently interpreted as both numeric and
categorical in different cycles. This happens because numeric
variables are identified by the presence of the phrase `"Range of
Values"` in the variables possible values. An example is available in

<https://github.com/cjendres1/nhanes/issues/21>


This is an initial attempt to find all such variables, and explore
possible ways forward.


# Demonstrate the problem

Known example:

```{r}
library(nhanesA)
```

Vanilla nhanes:

```{r}
nhanesOptions(use.db = FALSE)
str(nhanes("DEMO_G")$DMDHHSIZ)
str(nhanes("DEMO_H")$DMDHHSIZ)
```

Database version:

```{r}
nhanesOptions(use.db = TRUE)
str(nhanes("DEMO_G")$DMDHHSIZ)
str(nhanes("DEMO_H")$DMDHHSIZ)
```

FromURL version:

```{r}
str(nhanesFromURL("/Nchs/Nhanes/2011-2012/DEMO_G.XPT")$DMDHHSIZ)
str(nhanesFromURL("/Nchs/Nhanes/2013-2014/DEMO_H.XPT")$DMDHHSIZ)
```

# How often does this happen?

Possible options to find out are: 

* Do a variable summary of everything and find variables which are
  categorical in some but not all cycles.

* Use the codebooks stored in the database. Let's try the second one
  first because it will be faster.


```{r}
sql <- nhanesA:::.nhanesQuery
all_cb <- sql("select * from Metadata.VariableCodebook")
subset(all_cb, Variable == "DMDHHSIZ", select = 1:5)
```

So this 'finds' our known culprit.

Let's first restrict our attention to variables that are 'numeric' in
at least one table. There may be others, but we have very little hope
of finding them unless we inspect each manually (but see last section below).

```{r, warning = FALSE}
numeric_vars <- with(all_cb, unique(Variable[ValueDescription == "Range of Values"]))
numeric_cb <- subset(all_cb, Variable %in% numeric_vars, select = 1:5)
## quick check: which 'ValueDescription'-s look like numeric? Should be very few
maybe_numeric <- is.finite(as.numeric(numeric_cb$ValueDescription))
table(maybe_numeric)
```

We will focus on these variables.

```{r}
problem_vars <- unique(numeric_cb[maybe_numeric, ]$Variable)
str(problem_vars)
num_cb_byVar <- numeric_cb |> subset(Variable %in% problem_vars) |> split(~ Variable)
length(num_cb_byVar)
```

Let's start by summarizing these to keep only the unique (CodeOrValue,
ValueDescription) combinations, and then prioritize them by the number
of numeric-like values that remain.

```{r}
summary_byVar <-
    lapply(num_cb_byVar,
           function(d) unique(d[c("Variable", "CodeOrValue", "ValueDescription")]))
numNumeric <- function(d) suppressWarnings(sum(is.finite(as.numeric(d$ValueDescription))))
(nnum <- sapply(summary_byVar, numNumeric) |> sort())
```

The number of variables with 2 or more numeric variables seem like a
manageable number, so let's look at all of them.

```{r}
num_cb_byVar[ names(which(nnum > 2)) ]
num_cb_byVar[ names(which(nnum == 2)) ]
```

# What to do about these?

These all look like legitimate issues. Possible work-arounds are:

* Maintain an explicit list of such variables and handle them while
  creating the codebook. The least intrusive way would be to just
  insert a row with value description `"Range of
  Values"`. Additionally, we could drop the value descriptions which
  can be coerced to numeric.

* Maintain an explicit list of such variables and handle them in
  `nhanesTranslate()`

The advantage of doing this at the codebook level is that the database
will have the 'fixed' values.


# More aggressive conversion

A more aggressive strategy would be to do this whenever there is a
value description that can be coerced to a numeric value. This will
not require retaining a list of variables to fix, but will mean that
we may end up converting some variables to numeric even though it
didn't have `"Range of Values"` in any cycle.

To see the effect of this:

```{r}
categorical_cb <- subset(all_cb, !(Variable %in% numeric_vars), select = 1:5)
cat_num_combinations <- 
    unique(subset(categorical_cb, is.finite(as.numeric(ValueDescription)), select = 1:2))

for (i in 1:nrow(cat_num_combinations)) {
    v <- cat_num_combinations$Variable[[i]]
    t <- cat_num_combinations$TableName[[i]]
    try(cat(sprintf("---------\n%s > %s (%s)\n", t, v, nhanesCodebook(t)[[v]]$`SAS Label:` )))
    print(subset(categorical_cb, TableName == t & Variable == v))
}
```

The list, while fascinating in variety, does include definite false
positives, such as:

* `CDQ > CDQ070 (Sleep on 2+ pillows to help breathe)`
* `HUQ > HUQ050 (#Times received healthcare over past yr)`
* `OHXPRU_B > OHX14BPM (BOP: midfacial #14)`

Some of these could be added to our list, but detecting them
automatically seems problematic.


