---
title: "Initial Data Exploration of the `nhanesA` Package"
author: "Deepayan Sarkar"
output: github_document
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = "#>",
                      fig.path = "figures/",
                      out.width = "100%",
                      fig.width = 12, fig.height = 8,
                      dev = "svglite", dev.args = list(pointsize = 12),
                      cache = TRUE,
                      cache.path = "./knitr-cache/explore/")
options(width = 100)
library(lattice)
library(kableExtra) # for kable()
## library(DT)
source("R/utilities.R")
```


# Summary of available tables

Data in the `nhanesA` package are stored in a SQL Server database. The 
database can be queried using the (unexported) `.nhanesQuery()` function
or its public wrapper `phonto::nhanesQuery()`.

```{r}
library(nhanesA)
alltables <- nhanesA:::.nhanesQuery("select * from information_schema.tables;")
dim(alltables)
head(alltables) |> kable()
```

There are two versions of each table, the raw table with original codes, and 
translated versions with codes translated into human-readable values. In 
addition, there are some metadata and ontology tables.

```{r}
subset(alltables, !(TABLE_SCHEMA %in% c("Raw", "Translated"))) |> kable()
```

The translated table names can be extracted as follows.

```{r}
trtables <- subset(alltables, TABLE_SCHEMA == "Translated")$TABLE_NAME |> sort()
str(trtables)
```

Many of these start with `P_`; we will skip them (they represent releases made 
during the pandemic, and are inconsistent with previous releases). The remaining 
tables often have suffixes (separated by underscore) that indicate cycle, but 
may also have underscores in the main table name.

```{r}
std_tables <- trtables[!grepl("^P_", trtables)]
std_tables_split <- strsplit(std_tables, "_", fixed = TRUE)
std_tables[ sapply(std_tables_split, length) == 3 ]
```

We will assume that suffixes `_A`, `_B`, ..., `_J` indicate cycles. The number 
of times these prefixes appear are
```{r}
sapply(paste0("_", LETTERS[1:10]),
       function(s) sum(endsWith(std_tables, s)))
```


Removing these, we can get valid table names as follows.
```{r}
drop_table_suffix(std_tables) |> table() |> sort(decreasing = TRUE) |> head(100)
```

Some appear only once or twice. These are probably introduced later, and were not 
necessarily continued; for example, the following two tables relate to air quality.
```{r}
std_tables[startsWith(std_tables, "AQQ")]
nhanesCodebook('AQQ_E') |> lapply("[[", "SAS Label:") |> str()
```



# Using metadata tables to obtain table information

Alternatively, we can get information about available tables from the metadata 
tables, although for some reason, these do not include the `P_*` tables. 

```{r}
tableDesc <- nhanesA:::.nhanesQuery("select * from Metadata.QuestionnaireDescriptions")
tableDesc <- within(tableDesc, TableBase <- drop_table_suffix(TableName))
str(tableDesc)
```

We will work with  this from now on. We can summarize this table by table / questionnaire 
descriptions as follows.
```{r}
tableSummary <- 
  xtabs(~ TableBase + Description + DataGroup, tableDesc) |> 
  as.data.frame.table() |> subset(Freq > 0)
head(tableSummary, 20) |> kable() # use datatable() for html_output
```



# Extracting table data

Before trying to analyse any data, we may be interested in knowing whether the 
codebook has changed for a particular table across different cycles. Let's try 
this with the demographic tables.

```{r}
demotables <- get_table_names('DEMO', db = tableDesc)
cb.demo <- lapply(demotables, function(x) names(nhanesCodebook(x)))
str(cb.demo)
```

So the number of recorded variables keep changing. To combine across years, 
we will get all common variables, assuming that their interpretations have 
not changed over cycles.

```{r}
demovars <- get_common_vars(demotables)
str(demovars)
```

We can then merge all versions of a table by restricting to these common 
variables.

```{r}
demo.all <- merge_tables(demotables)
```

Some elementary summaries of the demographic variables:

```{r age_dist_yr}
qqmath(~ RIDAGEYR | SDDSRVYR, demo.all, plot.points = FALSE, distribution = qexp,
       f.value = ppoints(500), pch = ".", cex = 2, as.table = TRUE, grid = TRUE)
```

```{r race_dist_yr}
xtabs(~ SDDSRVYR + RIAGENDR + RIDRETH1, demo.all) |> 
  dotplot(auto.key = list(columns = 2), type = "o", 
          par.settings = simpleTheme(pch = 16))
```


# Combining tables for analysis

Suppose we want to combine demographic data with data from one or more other 
tables to perform some analysis. Let's start by looking at which tables have
been populated in most (at least 8) cycles. The `DEMO` table will not appear in 
this list because it has two slightly different descriptions (with frequencies 
7 and 3).

```{r}
subset(tableSummary, Freq > 7) |> kable()
```

Let's say we now want to combine the tables DEMO, WHQ (Weight history), BPX (Blood pressure), 
and BMX (Body measurements), which have the following variables.

```{r}
sapply(nhanesCodebook('WHQ'), "[[", "SAS Label:")
sapply(nhanesCodebook('BMX'), "[[", "SAS Label:")
sapply(nhanesCodebook('BPX'), "[[", "SAS Label:")
```

We can combine the tables individually as above.

```{r}
wtables <- c("DEMO", "WHQ", "BMX", "BPX")
tablist <- lapply(wtables, function(x) merge_tables(get_table_names(x, tableDesc)))
names(tablist) <- wtables
sapply(tablist, nrow)
```

We can now easily merge them by first subsetting to common `SEQN` values (in 
the same order).


```{r}
common_id <- Reduce(intersect, lapply(tablist, "[[", "SEQN"))
tablist_common <- lapply(tablist, function(d) d[match(common_id, d$SEQN), ])
dcombined <- Reduce(merge, tablist)
```

We are ready for some anaylsis. But variable names are still incomprehensible 
without referring to codebook. We can get variable descriptions from one of
the metadata tables.

```{r}
variableDesc <- nhanesA:::.nhanesQuery("select * from Metadata.QuestionnaireVariables")
```

This table does include the `P_*` tables, which we will exclude; because among 
other reasons, they have not been processed correctly, e.g.,

```{r}
subset(variableDesc, TableName |> startsWith("P_")) |> head()
```

even though <https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/P_ACQ.htm> has the 
relevant information. So, 

```{r}
variableDesc <- subset(variableDesc, !startsWith(TableName, "P_"))
uvarDesc <- subset(variableDesc, Variable %in% names(dcombined), select = c(Variable, SasLabel)) |> unique()
```

We can do a crude search over the SAS labels using `agrepl()`.

```{r}
subset(uvarDesc, agrepl("weight", SasLabel, ignore.case = TRUE, fixed = TRUE)) |> kable()
```

Now that we have a merged dataset, we can use it for standard data analysis.
Let's start by looking at how self-reported weight compares with actual 
measured weight.

```{r wt_comparison_1}
xyplot(WHD020 ~ BMXWT, dcombined)
```

The reason for the weird values on the y-axis is that special codes for `WHD020`
have not been translated. 
```{r}
nhanesCodebook("BMX")[["BMXWT"]][["BMXWT"]] # no issues - only missing
subset(variableDesc, Variable == "WHD020", select = c(1, 2, 5)) |> kable()
nhanesCodebook("WHQ")[["WHD020"]][["WHD020"]] # 77777 / 99999
nhanesCodebook("WHQ_B")[["WHD020"]][["WHD020"]] # 7777 / 9999
```

It is not clear that there is a good solution to this problem, but translating
these to `NA`s would probably be better than doing nothing.

After dropping unrealistically high weights, we get a more reasonable plot.

```{r wt_comparison_2}
xyplot(I(WHD020 * 0.453592) ~ BMXWT | RIAGENDR, dcombined, subset = WHD020 < 1000, 
       alpha = 0.25, pch = ".", cex = 3, 
       abline = c(0, 1), smooth = "lm", grid = TRUE, aspect = "iso")
```

The next plot explores whether average (measured) weight varies with age.

```{r wt_by_age}
xyplot(BMXWT ~  RIDAGEYR | RIAGENDR, dcombined,
       alpha = 0.25, pch = ".", cex = 3, smooth = "loess", 
       grid = TRUE, col.line = 1)
```


# Variables and tables they belong to

Unfortunately, this scheme of merging tables is fundamentally flawed for one important 
reason: the same variable may not be recorded in the same table in all cycles.
Consider tables whose description contains "cholesterol". 

```{r}
subset(tableSummary, grepl("cholesterol", Description, ignore.case = TRUE)) |> kable()
```

Clearly, similar data  is included in tables with different names. This becomes 
more obvious when we use the variable metadata table to look up a particular variable
of interest. Let's first look at the variables in one of the `TRIGLY` datasets 
to pick one that could be of interest.

```{r}
subset(tableDesc, TableBase == "TRIGLY")[1:4] |> kable()
sapply(nhanesCodebook("TRIGLY_E"), "[[", "SAS Label:")
```

`LBDLDL` seems interesting, so let's search for the  variable in all tables.

```{r}
subset(variableDesc, Variable == "LBDLDL")[c(1, 2, 3)] |> kable()
```

Upshot: To combine different variables of interest in preparation for data 
analysis, combining datasets is not the correct approach. Instead, we will 
need to select specific variables of interest, and then collect them from 
whichever table they happen to be available in.

# Miscellaneous oddities

## Table name mismatch

There is a mismatch for tables with base name `SSDFS` (there may be others as 
well).

```{r}
std_tables[startsWith(std_tables, "SSDFS")]
subset(tableDesc, startsWith(TableName, "SSDFS"))[1:5]
```

The `SSDFS_A` table does exist:
```{r}
str(nhanes('SSDFS_A', translated = TRUE))
```
However, even though `SSDFS_G` is a legitimate table, there is no <https://wwwn.cdc.gov/Nchs/Nhanes/2011-2012/SSDFS_A.htm>, and 
the corresponding codebook naturally has no useful information. Where did 
the `SSDFS_A` table come from?



## Should any tables end with `_A`?

Somewhat related to the above: The following tables end with `_A`, but there 
should not be any, because tables in the first did not have any cycle suffix.

```{r}
(tables_with_A <- std_tables[ endsWith(std_tables, "_A") ])
```

Let's check if they are available:

```{r}
sapply(tables_with_A, nhanes) |> sapply(FUN = nrow)
```

Let's check if the version without the suffix are available:

```{r}
sapply(gsub("_A$", "", tables_with_A), 
       function(x) {
         e <- try(nhanes(x), silent = TRUE)
         if (inherits(e, "try-error")) NA_real_ else nrow(e)
       })
```

So all these seem legitimate. In fact, they do have websites, but with some
unexpected twists, e.g.

- <https://wwwn.cdc.gov/Nchs/Nhanes/1999-2000/L02HPA_A.htm> 
  (Hepatitis-A, which may or may not be where the A suffix comes from)

- <https://wwwn.cdc.gov/Nchs/Nhanes/1999-2000/SSAFB_A.htm>
  (originally published 2012, but revised 2022 to correct for some procedural 
  errors. This seems to be a common theme in most of these examples)

- <https://wwwn.cdc.gov/Nchs/Nhanes/1999-2000/SSDFS_A.htm>
  (this one uses lefotver sera from participants in the first cycle, but
  was actually published later in 2022)

Someone should look at these systematically, but apart from the first one 
(L02HPA), these seem legitimate.

For the first one, we have

```{r}
std_tables[ startsWith(std_tables, "L02HPA")]
```

Looking more closely, the actual filename for the first cycle is `L02HPA_a.xpt`, 
with the lowercase `_a` suffix. The subsequent cycles should have been 
`L02HPA_a_B.xpt`, `L02HPA_a_C.xpt`, etc., but of course there were not.

Anyway, so all this looks OK. _However_, the metadata table doesn't always 
track these consistently, e.g.,

```{r}
subset(tableDesc, startsWith(TableName, "L02HPA"))
```

has missing documentation / file links for the first one (because of the 
lowercase `_a`?).


## Bug? `includelabels = TRUE` does not seem to do anything

What is the difference between `includelabels = TRUE` and `FALSE`? Neither
version seems to include the SAS labels


```{r}
nhanes('DEMO_G', includelabels = FALSE) |> attributes() |> str()
nhanes('DEMO_G', includelabels = TRUE) |> attributes() |> str()
```


## Potentially missing tables

Finally, just a sanity check to verify that all tables that should be are 
in fact included in the metadata table.

```{r}
all_tables_in_db <- 
  subset(nhanesA:::.nhanesQuery("select * from information_schema.tables;"),
         TABLE_SCHEMA == "Translated")$TABLE_NAME
all_tables_in_metadata <- 
  nhanesA:::.nhanesQuery("select * from Metadata.QuestionnaireDescriptions")$TableName
anyDuplicated(all_tables_in_db)         # none
anyDuplicated(all_tables_in_metadata)   # none
setdiff(all_tables_in_metadata, all_tables_in_db) # none
setdiff(all_tables_in_db, all_tables_in_metadata) |> 
  grep(pattern = "^P_", invert = TRUE, value = TRUE)
setdiff(all_tables_in_db, all_tables_in_metadata) |> 
  grep(pattern = "^P_", invert = FALSE, value = TRUE)
```

