---
title: "Frequency of NHANES releases"
author: "Deepayan Sarkar"
output: github_document
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = "#>",
                      fig.path = "figures/release-freq-",
                      out.width = "100%",
                      fig.width = 12, fig.height = 8,
                      dev = "svglite", dev.args = list(pointsize = 12),
                      cache = TRUE,
                      cache.path = "./knitr-cache/relfreq/")
options(width = 100)
library(lattice)
```


# How often are NHANES data released / updated

```{r}
library(nhanesA)
manifest <- nhanesManifest()
```

We want to tabulate when NHANES releases data. We expect that bulk
releases of tables happen all together, while some tables may be
released or updated on a as-needed basis.

The release information is available in the manifest as the
`Date.Published` field:
```r
xtabs(~ Date.Published, manifest) |> sort() |> tail(20)
```

So let's try to parse this systematically.

```{r}
pubdate <- manifest$Date.Published
updates <- startsWith(pubdate, "Updated")
datesplit <- strsplit(pubdate, split = "[[:space:]]")
mismatch <- (sapply(datesplit, length) - updates) != 2L
if (any(mismatch)) {
    cat("Unexpected format:\n")
    print(manifest[mismatch, , drop = FALSE])
}
```
This currently flags one withdrawn table (PAHS_H). We will just skip this and proceed.

```{r}
pubdate <- pubdate[!mismatch]
updates <- updates[!mismatch]
datesplit <- datesplit[!mismatch]
datesplit[updates] <- lapply(datesplit[updates], "[", -1)
print(table(sapply(datesplit, length))) # should be all 2
pub_summary <-
    data.frame(updated = updates,
               year = sapply(datesplit, "[[", 2) |> as.numeric(),
               month = sapply(datesplit, "[[", 1) |> factor(levels = month.name))
```

Plot frequency by year:

```{r, byyear}
xtabs(~ year + updated, pub_summary) |>
    barchart(horizontal = FALSE, ylab = "Number of tables",
             auto.key = list(text = c("Original", "Update"), columns = 2),
             scales = list(x = list(rot = 45)))
```


Table of frequency by month:

```{r}
pubfreq <- xtabs(~ interaction(month, year, sep = "-") + updated, pub_summary)
pubfreq0 <- pubfreq[rowSums(pubfreq) > 0, , drop = FALSE]
pubfreq0
```

There are a bit too many months to plot nicely, but approximately:

```{r, bymonth}
npub <- rowSums(pubfreq)
npub.date <- as.Date(paste0("01", "-", names(npub)), format = "%d-%B-%Y")
xyplot(npub ~ npub.date, type = "h", grid = TRUE,
       xlab = "Month", ylab = "Number of tables published / updated") +
    latticeExtra::layer(panel.text(x[y > 30], y[y > 30],
                                   format(x[y > 30], "%Y-%m"),
                                   pos = 3, cex = 0.75))
```


