---
title: "Potential Issues with `nhanesA` and `phonto`"
author: "Deepayan Sarkar"
output: github_document
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = "# ",
                      fig.path = "figures/",
                      out.width = "100%",
                      fig.width = 12, fig.height = 5,
                      dev = "svglite", dev.args = list(pointsize = 12),
                      cache = TRUE,
                      cache.path = "./knitr-cache/issues/")
options(width = 100)
library(lattice)
library(kableExtra) # for kable()
## library(DT)
library(nhanesA)
library(phonto)
source("R/utilities.R")
```


# List of available tables

Data in the `nhanesA` package (or more precisely, in the CCB `NHANES` Docker) 
are stored in a SQL Server database. The  database can be queried using the 
(unexported) `nhanesA:::.nhanesQuery()` function or its public wrapper 
`phonto::nhanesQuery()`.

There is no 'R-like' way (that I can tell) to get a list of all available 
tables. However, there are at least two ways to get this via direct SQL queries:

```{r}
all_tables_in_db <- 
  subset(nhanesA:::.nhanesQuery("select * from information_schema.tables;"),
         TABLE_SCHEMA == "Translated")$TABLE_NAME
all_tables_in_metadata <- 
  nhanesA:::.nhanesQuery("select * from Metadata.QuestionnaireDescriptions")$TableName
```

The second one uses one of the several metadata tables stored in the database.
The first one is the list of actual tables available (labeled as "Translated").
As a sanity check, we can verify that the list of "Raw" and "Translated" tables 
is identical.
```{r}
all_raw_tables_in_db <- 
  subset(nhanesA:::.nhanesQuery("select * from information_schema.tables;"),
         TABLE_SCHEMA == "Raw")$TABLE_NAME
identical(sort(all_tables_in_db), sort(all_raw_tables_in_db))
```

Also, all the tables listed in the metadata table are actually available:
```{r}
setdiff(all_tables_in_metadata, all_tables_in_db)
```

However, the converse is not true: not all translated tables in the database 
are listed in the metadata. These are of two types: those starting with `P_`
and the rest, listed separately below:

```{r}
setdiff(all_tables_in_db, all_tables_in_metadata) |> 
  grep(pattern = "^P_", invert = FALSE, value = TRUE)
```
(these are intentionally excluded --- we may or may not want to revisit that) 
and the rest:
```{r}
setdiff(all_tables_in_db, all_tables_in_metadata) |> 
  grep(pattern = "^P_", invert = TRUE, value = TRUE)
```
Most of these seem to be legitimate tables, so they should not be excluded. 
Here are corresponding links to the NHANES website for easy exploration.
```{r}
URLs <- 
  setdiff(all_tables_in_db, all_tables_in_metadata) |> 
  grep(pattern = "^P_", invert = TRUE, value = TRUE) |>
  sapply(FUN = nhanes_url)
```

```{r results="asis", comment = ""}
cat("", sprintf("- [%s](%s)", names(URLs), URLs), sep = "\n")
```

All of these tables actually exist in the database; but they appear not to 
have proper codebooks. For example,
```{r}
nhanes("POOLTF_E") |> str()
nhanesCodebook("POOLTF_E") |> str()
```
This one is particularly interesting, because for the next cycle, we have
```{r}
nhanesCodebook("POOLTF_F") |> str()
```
Looking at the corresponding webpages, it turns out that `POOLTF_D` and 
`POOLTF_E` don't have a proper codebook, so that's probably what is causing 
the failure. Several others (but not all) have similar problems.

These two have the problem that they don't have the proper suffix, so their URL 
cannot be computed from the table names:

- <https://wwwn.cdc.gov/Nchs/Nhanes/2001-2002/PFC_POOL.htm>

- <https://wwwn.cdc.gov/nchs/nhanes/2001-2002/SSNH4THY.htm>


# Metadata for `L02HPA_A`

This may be a completely one-off problem in the table metadata:

```{r}
tableDesc <- nhanesA:::.nhanesQuery("select * from Metadata.QuestionnaireDescriptions")
subset(tableDesc, TableName |> startsWith("L02HPA"))
```

Note that the `DocFile` and `DataFile` entries are empty for the first of these, 
even though this link works:
```{r results="asis"}
cat("\n\n<", nhanes_url("L02HPA_A"), ">\n\n", sep = "")
```

The one unusual thing about this table is that the `_A` suffix was not actually
a cycle indicator (and it was originally a lowercase `_a`), though the original 
meaning was subsequently dropped.


# Things that would be nice to have

(Maybe they already exist)

- `browseNHANES(nh_table = "BPQ_J")` etc. opens NHANES website --- maybe we can 
  add the option to return the URL instead of opening it (could be helpful when 
  creating our own summary), or have a different `getNHANES_URL()` with the 
  same arguments. I have a limited implementation `nhanes_url()` in 
  [this file](R/utilities.R).

- `nhanesAttr()` could return number of `NA` values columnwise (currently 
  returns total)

```{r}
nhanesAttr("BPQ_J") |> str()
```


- Searchable table of NHANES tables (using datatable maybe). Something better than 
  <https://wwwn.cdc.gov/nchs/nhanes/search/DataPage.aspx>. Either have 
  pre-built versions, or generate on-the-fly, depending on speed. I am 
  planning to work on this anyway as it will be useful for downstream 
  analysis, but we can discuss whether it would be something useful to have
  either in `nhanesA` or `phonto`.

- Some way to navigate between tables and variables --- in particular, it's 
  perhaps not obvious that the same variable can be in different tables in 
  different years, and even in the same year. Note that there is already 
  `nhanesSearchTableNames()` and `nhanesSearchVarName()`, e.g.,
  
```{r}
nhanesSearchVarName('PHAFSTMN')
```



# Example code

The output produced is not what I would expect

- `?dataDescription`

```{r}
dataDescription(list(BPQ_J=c("BPQ020", "BPQ050A"), DEMO_J=c("RIDAGEYR","RIAGENDR")))
```

- `?nhanesSearchTableNames` - also doesn't set `details = TRUE` as documented

```{r}
hepbd <- nhanesSearchTableNames('HEPBD', includeurl=TRUE)
dim(hepbd)
str(hepbd)
```

- `?nhanes` - No visible difference. Bug?

```{r}
nhanes('DEMO_G', includelabels = FALSE) |> attributes() |> str()
nhanes('DEMO_G', includelabels = TRUE) |> attributes() |> str()
```



# Some curiosities

These are not problems exactly, but maybe deserve some thought.

One reason for translating tables is to make data comparable across cycles.
The task of translation is complicated when a _continuous_ measurement uses
specific numeric values to encode special meaning. It is not clear how these
can be handled without introducing new variables, but not handling this may
lead to user confusion.

Not surprisingly, the special coding may even change from cycle to cycle. For 
example,
```{r}
nhanesCodebook("WHQ")[["WHD020"]][["WHD020"]] # 77777 / 99999
nhanesCodebook("WHQ_B")[["WHD020"]][["WHD020"]] # 7777 / 9999
```
A not-too-difficult solution is to convert all special values to missing, 
though that will lead to loss of information.


Finally, is the CRAN repository intentionally fixed to an old snapshot?
This could be a Bioconductor-like design, which is OK, but it's not what one
normally expects from CRAN.

```{r}
R.version[["version.string"]]
getOption("repos")
```

The relevant parts in the Docker file that controls these are:

```sh
ENV R_VERSION_MAJOR 4
ENV R_VERSION_MINOR 2
ENV R_VERSION_BUGFIX 1
ENV R_REPOSITORY=https://packagemanager.posit.co/cran/__linux__/focal/2022-10-18
```

Also the base image from 2019

```sh
FROM mcr.microsoft.com/mssql/server:2019-CU12-ubuntu-20.04
```

seems to have a [2022 update](https://hub.docker.com/_/microsoft-mssql-server) 
(with the same version of Ubuntu though).

