---
title: "Variables in NHANES"
author: "Deepayan Sarkar"
output: github_document
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = "#>",
                      fig.path = "figures/",
                      out.width = "100%",
                      fig.width = 12, fig.height = 5,
                      dev = "svglite", dev.args = list(pointsize = 12),
                      cache = TRUE,
                      cache.path = "./knitr-cache/variables/")
options(width = 100)
library(kableExtra) # for kable()
## library(DT)
source("R/utilities.R")
sort.data.frame <- function(x, by, decreasing = FALSE, ...) {
  f <- eval(attr(stats::terms(by), "variables"), x, environment(by))
  o <- do.call(order, c(f, list(decreasing = decreasing, ...)))
  x[o, , drop = FALSE]
}
```


# Summary of available variables

We can get information about available variables from the metadata 
table `Metadata.QuestionnaireVariables`. 

```{r}
library(nhanesA)
variableDesc <- nhanesA:::.nhanesQuery("select * from Metadata.QuestionnaireVariables")
dim(variableDesc)
```

There are not actually these many variables, because every occurrence in every 
table is recorded separately.

```{r}
length(unique(variableDesc$Variable))
```

Let's start by dropping the `P_*` tables for now.

```{r}
variableDesc <- subset(variableDesc, !startsWith(TableName, "P_"))
dim(variableDesc)
```

Most "useful" variables would be recorded in most (if not all) cycles; so let's 
see how many appear in at least 7 tables.

```{r}
table(variableDesc$Variable) |> sort(decreasing = TRUE) |> table()
```

What does it mean for a variable to appear in more than 10 tables? This makes 
sense for `SEQN`:

```{r}
sum(variableDesc$Variable == "SEQN")
```

But what are the other ones?

```{r}
variableFreq <- table(variableDesc$Variable) |> sort(decreasing = TRUE)
variableFreq[variableFreq > 13]
```

Let's look at one of these.

```{r}
subset(variableDesc, Variable == "PHAFSTMN", select = c(1, 2, 5)) #|> kable()
```

This raises the obvious question: Are these measuring the same thing? 


# Variables defined in multiple tables

Let's look at actual data in `FASTQX_D`, `GLU_D`, and `OGTT_D`.

```{r}
wtables <- c("FASTQX_D", "GLU_D", "OGTT_D")
tablist <- sapply(wtables, 
                  function(name) nhanes(name)[c("SEQN", "PHAFSTMN")], 
                  simplify = FALSE)
str(tablist)
```

Let's check if they are identical for every common `SEQN`:

```{r}
str(with(tablist, sort(intersect(FASTQX_D$SEQN, GLU_D$SEQN))))
str(with(tablist, sort(intersect(FASTQX_D$SEQN, OGTT_D$SEQN))))
str(with(tablist, sort(intersect(GLU_D$SEQN, OGTT_D$SEQN))))
keep <- with(tablist, 
             FASTQX_D$SEQN |> intersect(GLU_D$SEQN) |> intersect(OGTT_D$SEQN))
keep <- sort(keep) # SEQN to get from each table
keep.index <- lapply(tablist, function(d) match(keep, d$SEQN) )
sublist <- sapply(wtables, 
                  function(name) tablist[[name]][keep.index[[name]] , ],
                  simplify = FALSE)
all.equal(sublist$FASTQX_D, sublist$GLU_D, check.attributes = FALSE)
all.equal(sublist$FASTQX_D, sublist$OGTT_D, check.attributes = FALSE)
```

So at least there seems to be some consistency.

# Variable names and descriptions

Ideally, each variable should have the same description and SAS Label whenever
it appears. Let's look at unique combinations.

```{r}
unique(variableDesc$Variable) |> length()
unique(variableDesc[c("Variable", "Description", "SasLabel")]) |> dim()
unique(variableDesc[c("Variable", "Description")]) |> dim()
unique(variableDesc[c("Variable", "SasLabel")]) |> dim()
```

So not too bad. Let's look at the variables that have multiple `SasLabels` first.

```{r}
multLabels <- subset(unique(variableDesc[c("Variable", "SasLabel")]),
                     Variable %in% names(which(table(Variable) > 1)))
sort(multLabels, by = ~ Variable) |> head(100) |> kable()
```

This is mostly OK. But we need to look into some variables which have potentially
different units of measurement.

```{r}
subset(multLabels, endsWith(SasLabel, "g)")) |>
  subset(Variable %in% names(which(table(Variable) > 1))) |>
  sort(by = ~ Variable) |> kable()
```

Similarly those with multiple descriptions.

```{r}
multDesc <- subset(unique(variableDesc[c("Variable", "Description")]),
                   Variable %in% names(which(table(Variable) > 1)))
sort(multDesc, by = ~ Variable) |> head(100) |> kable()
```

# Summary

Subject to some caveats (consistency in units, potentially conflicting data in 
different tables for same participant), a variable centric workflow seems 
feasible, once we have a good way to select relevant variables and deal with 
missing data.

The following exports the unique combinations of variable names, labels, and 
descriptions for exploration using other tools.

```{r}
write.csv(unique(variableDesc[c("Variable", "Description", "SasLabel")]) |> 
            sort(by = ~ Variable),
          file = "nhanes-variables.csv", row.names = FALSE)
```


