---
title: "Variables in NHANES"
author: "Deepayan Sarkar"
output: github_document
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = "#>",
                      fig.path = "figures/",
                      out.width = "100%",
                      fig.width = 12, fig.height = 5,
                      dev = "svglite", dev.args = list(pointsize = 12),
                      cache = TRUE,
                      cache.path = "./knitr-cache/variables/")
options(width = 100)
library(kableExtra) # for kable()
## library(DT)
source("R/utilities.R")
sort.data.frame <- function(x, by, decreasing = FALSE, ...) {
  f <- eval(attr(stats::terms(by), "variables"), x, environment(by))
  o <- do.call(order, c(f, list(decreasing = decreasing, ...)))
  x[o, , drop = FALSE]
}
```


# Summary of available variables

We can get information about available variables from the metadata 
table `Metadata.QuestionnaireVariables`. 

```{r}
library(nhanesA)
variableDesc <- nhanesA:::.nhanesQuery("select * from Metadata.QuestionnaireVariables")
dim(variableDesc)
```

There are not actually these many variables, because every occurrence in every 
table is recorded separately.

```{r}
length(unique(variableDesc$Variable))
```

Let's start by dropping the `P_*` tables for now.

```{r}
variableDesc <- subset(variableDesc, !startsWith(TableName, "P_"))
dim(variableDesc)
```

Most "useful" variables would be recorded in most (if not all) cycles; so let's 
see how many appear in at least 7 tables.

```{r}
table(variableDesc$Variable) |> sort(decreasing = TRUE) |> table()
```

What does it mean for a variable to appear in more than 10 tables? This makes 
sense for `SEQN`:

```{r}
sum(variableDesc$Variable == "SEQN")
```

But what are the other ones?

```{r}
variableFreq <- table(variableDesc$Variable) |> sort(decreasing = TRUE)
variableFreq[variableFreq > 13]
```

Let's look at one of these.

```{r}
subset(variableDesc, Variable == "PHAFSTMN", select = c(1, 2, 5)) #|> kable()
```

This raises the obvious question: Are these measuring the same thing? 


# Variables defined in multiple tables

Let's look at actual data in `FASTQX_D`, `GLU_D`, and `OGTT_D`.

```{r}
wtables <- c("FASTQX_D", "GLU_D", "OGTT_D")
tablist <- sapply(wtables, 
                  function(name) nhanes(name)[c("SEQN", "PHAFSTMN")], 
                  simplify = FALSE)
str(tablist)
```

Let's check if they are identical for every common `SEQN`:

```{r}
str(with(tablist, sort(intersect(FASTQX_D$SEQN, GLU_D$SEQN))))
str(with(tablist, sort(intersect(FASTQX_D$SEQN, OGTT_D$SEQN))))
str(with(tablist, sort(intersect(GLU_D$SEQN, OGTT_D$SEQN))))
keep <- with(tablist, 
             FASTQX_D$SEQN |> intersect(GLU_D$SEQN) |> intersect(OGTT_D$SEQN))
keep <- sort(keep) # SEQN to get from each table
keep.index <- lapply(tablist, function(d) match(keep, d$SEQN) )
sublist <- sapply(wtables, 
                  function(name) tablist[[name]][keep.index[[name]] , ],
                  simplify = FALSE)
all.equal(sublist$FASTQX_D, sublist$GLU_D, check.attributes = FALSE)
all.equal(sublist$FASTQX_D, sublist$OGTT_D, check.attributes = FALSE)
```

So at least there seems to be some consistency.

# Variable names and descriptions

Ideally, each variable should have the same description and SAS Label whenever
it appears. Let's look at unique combinations.

```{r}
unique(variableDesc$Variable) |> length()
unique(variableDesc[c("Variable", "Description", "SasLabel")]) |> dim()
unique(variableDesc[c("Variable", "Description")]) |> dim()
unique(variableDesc[c("Variable", "SasLabel")]) |> dim()
```

So not too bad. Let's look at the variables that have multiple `SasLabels` first.

```{r}
multLabels <- subset(unique(variableDesc[c("Variable", "SasLabel")]),
                     Variable %in% names(which(table(Variable) > 1)))
sort(multLabels, by = ~ Variable) |> head(100) |> kable()
```

This is mostly OK. But we need to look into some variables which have potentially
different units of measurement.

```{r}
subset(multLabels, endsWith(SasLabel, "g)")) |>
  subset(Variable %in% names(which(table(Variable) > 1))) |>
  sort(by = ~ Variable) |> kable()
```

This is a real issue in at least one example, where the data are from pooled 
samples, hence identified by `SAMPLEID`, not `SEQN` (see below).
```{r}
source("R/variable-extraction.R")
lbc028 <- get_variable_data("LBC028", db, idvar = "SAMPLEID", verbose = FALSE)
with(lbc028, tapply(LBC028, TABLE, quantile, na.rm = TRUE)) |> do.call(what = rbind) |> kable()
```

This one seems fine:
```{r}
lbxsossi <- get_variable_data("LBXSOSSI", db, idvar = "SEQN", verbose = FALSE)
with(lbxsossi, tapply(LBXSOSSI, TABLE, quantile, na.rm = TRUE)) |> do.call(what = rbind)# |> kable()
```


Similarly, the following are variables with multiple descriptions.

```{r}
multDesc <- subset(unique(variableDesc[c("Variable", "Description")]),
                   Variable %in% names(which(table(Variable) > 1)))
sort(multDesc, by = ~ Variable) |> head(100) |> kable()
```

# Summary

Subject to some caveats (consistency in units, potentially conflicting data in 
different tables for same participant), a variable centric workflow seems 
feasible, once we have a good way to select relevant variables and deal with 
missing data.

The following exports the unique combinations of variable names, labels, and 
descriptions for exploration using other tools.

```{r}
write.csv(unique(variableDesc[c("Variable", "Description", "SasLabel")]) |> 
            sort(by = ~ Variable),
          file = "nhanes-variables.csv", row.names = FALSE)
```


Actually, to be useful, this also needs to record how many unique participants
each variable has records for, because many variables are very similar, but with
substantially different coverage. This will require actually loading and 
inspecting the relevant datasets.

# Participant ID

Participants in different tables are matched by `SEQN`. However, not all tables 
have a `SEQN` variable. Some tables report pooled data, and have a `SAMPLEID` 
variable identifying the pooled sample. We can check which tables have which 
variable using the following code, but it takes a while to run, so we will 
not run it.
```r
all_tables_in_metadata <- 
  nhanesA:::.nhanesQuery("select * from Metadata.QuestionnaireDescriptions")$TableName
idvars <- 
  sapply(all_tables_in_metadata,
         function(tab) { 
           cat("\r", tab, ".....")
           c("SEQN", "SAMPLEID") %in% names(nhanes(tab))
         })
rownames(idvars) <- c("SEQN", "SAMPLEID")
```

Of these, the ones potentially of interest are those that do _not_ have `SEQN`, 
as well as those that have `SAMPLEID`. These are given by
```r
unusual_tables <- subset(as.data.frame(t(idvars)), !SEQN | SAMPLEID)
```
which we will define manually:
```{r}
unusual_tables <- 
  structure(list(
    SEQN = c(FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, 
             FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, 
             FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, 
             TRUE, TRUE, TRUE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, 
             FALSE, FALSE, FALSE, FALSE, FALSE, TRUE), 
    SAMPLEID = c(TRUE, TRUE, TRUE, TRUE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, 
                 FALSE, FALSE, TRUE, TRUE, TRUE, TRUE, FALSE, FALSE, TRUE, TRUE, 
                 TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, FALSE, FALSE, 
                 FALSE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE)
    ), 
    row.names = c("BFRPOL_E", "BFRPOL_F", "BFRPOL_G", "BFRPOL_H", "DRXFCD_C", "DRXFCD_D", "DRXFCD_F", 
                  "DRXFCD_G", "DRXFCD_H", "DRXMCD_C", "DRXMCD_D", "DRXMCD_E", "DRXMCD_F", 
                  "PCBPOL_D", "PCBPOL_E", "PCBPOL_G", "PCBPOL_I", "SSPCB_B", "SSPST_B", 
                  "PSTPOL_E", "PSTPOL_F", "PSTPOL_I", "PSTPOL_H", "DOXPOL_E", "POOLTF_G", 
                  "POOLTF_H", "POOLTF_I", "BFRPOL_D", "BFRPOL_I", "SSBFR_B", "DRXFCD_E", 
                  "DRXMCD_G", "PCBPOL_F", "PCBPOL_H", "PSTPOL_D", "PSTPOL_G", "DOXPOL_D", 
                  "DOXPOL_F", "DOXPOL_G", "POOLTF_F"), class = "data.frame")
```

The following tables have both:

```{r}
subset(unusual_tables, SEQN & SAMPLEID) |> kable()
```

These define how individuals are pooled into samples, e.g., see
<https://wwwn.cdc.gov/Nchs/Nhanes/2011-2012/POOLTF_G.htm>. 

_NOTE_ however that `SAMPLEID` is only unique within a cycle, unlike `SEQN`, 
so they must be matched to participants in their own cycle.

The following tables have `SAMPLEID` but not `SEQN`; these are presumably the 
results of analysis done on pooled samples.

```{r}
subset(unusual_tables, !SEQN & SAMPLEID) |> kable()
```

This are in fact all the tables that have the string `POL_` in them:

```{r}
all_tables_in_metadata <- 
  nhanesA:::.nhanesQuery("select * from Metadata.QuestionnaireDescriptions")$TableName
identical(sort(grep("POL_", all_tables_in_metadata, value = TRUE)),
          sort(rownames(subset(unusual_tables, !SEQN & SAMPLEID))))
```

Finally, the following tables have neither `SAMPLEID` nor `SEQN`.
```{r}
subset(unusual_tables, !SEQN & !SAMPLEID) |> kable()
```

These appear to be

- `DRXFCD` : Dietary Interview Technical Support File - Food Codes
- `DRXMCD` : Dietary Interview Technical Support File - Modification Codes

`SS(PCB|PST|BFR)_B` seem to refer to a different set of pooling; see
<https://wwwn.cdc.gov/nchs/nhanes/2001-2002/SSPCB_B.htm>. Here the ID variable
is `POOLID`. This variable only appears in these three tables.
```{r}
subset(variableDesc, Variable == "POOLID")[c(1, 2, 5)] |> kable()
```
So these three may be analysed together, but most likely cannot be combined 
with any others.

# Tables where participant ID is repeated

It would be convenient to assume that `SEQN` is unique (for the majority of 
tables where it is present), but this is not true for all tables; e.g., see

<https://wwwn.cdc.gov/Nchs/Nhanes/2003-2004/dr2iff_c.htm>

We can check which tables have duplicated `SEQN` as follow, but again we will 
not run this code as it is slow.

```r
id_repeated <- 
  sapply(sort(all_tables_in_metadata),
         function(tab) {
           cat("\r", tab, ".......")
           d <- nhanes(tab)
           !is.null(d[["SEQN"]]) && !(anyDuplicated(d[["SEQN"]]) == 0)
         })
which_repeated <- names(id_repeated)[id_repeated]
```

The result is 

```{r}
which_repeated <- 
  c("AUXAR_I",  "AUXAR_J",  "AUXTYM_I", "AUXTYM_J", "AUXWBR_I", "AUXWBR_J", 
    "DR1IFF_C", "DR1IFF_D", "DR1IFF_E", "DR1IFF_F", "DR1IFF_G", "DR1IFF_H", 
    "DR1IFF_I", "DR1IFF_J", "DR2IFF_C", "DR2IFF_D", "DR2IFF_E", "DR2IFF_F", 
    "DR2IFF_G", "DR2IFF_H", "DR2IFF_I", "DR2IFF_J", "DRXIFF", "DRXIFF_B", 
    "DS1IDS_E", "DS1IDS_F", "DS1IDS_G", "DS1IDS_H", "DS1IDS_I", "DS1IDS_J", 
    "DS2IDS_E", "DS2IDS_F", "DS2IDS_G", "DS2IDS_H", "DS2IDS_I", "DS2IDS_J", 
    "DSQ2_B",   "DSQ2_C",   "DSQ2_D",   "DSQFILE2", "DSQIDS_E", "DSQIDS_F", 
    "DSQIDS_G", "DSQIDS_H", "DSQIDS_I", "DSQIDS_J", "FFQDC_C", "FFQDC_D", 
    "PAQIAF",   "PAQIAF_B", "PAQIAF_C", "PAQIAF_D", "PAXDAY_G", "PAXDAY_H", 
    "PAXHR_G",  "PAXHR_H",  "RXQ_ANA",  "RXQ_RX",   "RXQ_RX_B", "RXQ_RX_C", 
    "RXQ_RX_D", "RXQ_RX_E", "RXQ_RX_F", "RXQ_RX_G", "RXQ_RX_H", "RXQ_RX_I", 
    "RXQ_RX_J", "RXQANA_B", "RXQANA_C", "SSHPV_F")
```



