---
title: "Creating a Searchable Table of Variables in NHANES"
author: "Deepayan Sarkar"
output: github_document
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = "#>",
                      out.width = "100%",
                      cache = TRUE,
                      cache.path = "./knitr-cache/variable-search/")
options(width = 100)
library(kableExtra) # for kable()
library(DT)
sort.data.frame <- function(x, decreasing = FALSE, by = NULL, ...)
{
    if (!inherits(by, "formula")) stop("'by' must be a formula")
    f <- .formula2varlist(by, x)
    o <- do.call(order, c(unname(f), list(decreasing = decreasing, ...)))
    x[o, , drop = FALSE]
}
```

# Summary of available variables

The NHANES data consist of multiple tables over multiple cycles, each
with a set of recorded variables. The variable names themselves are
fairly cryptic, and not useful as search terms. Fortunately, the
NHANES website also provides descriptions of these variables in
comprehensive tables (grouped by components such as
[Demographics](https://wwwn.cdc.gov/nchs/nhanes/search/variablelist.aspx?Component=Demographics),
[Laboratory](https://wwwn.cdc.gov/nchs/nhanes/search/variablelist.aspx?Component=Laboratory),
etc.), along with a search interface.


The nhanesA package can download these tables and make the information
contained in them available as a data frame using the
`nhanesManifest()` function.

```{r}
library(nhanesA)
varmf <- nhanesManifest("variables")
```

This downloads and combined several large web pages, so it may take a
little time to run. Add `verbose = TRUE` to see some indication of
what is happening. 

The first few rows of the resulting data frame are given by

```{r}
head(varmf)
```

and it dimensions are

```{r}
dim(varmf)
```

If we drop the pandemic (`P_*`) tables and limited access tables, we get
```{r}
varmf <- subset(varmf, !startsWith(Table, "P_") & UseConstraints == "None")
dim(varmf)
```

There are not actually these many distinct variables, because every
occurrence in every table is recorded separately.

```{r}
length(unique(varmf$VarName))
```

We may be more interested in variables that are recorded in most (if
not all) cycles. Let's see how frequently variables appear.

```{r}
table(varmf$VarName) |> sort(decreasing = TRUE) |> table()
```

What does it mean for a variable to appear in more than 10 tables? This makes 
sense for `SEQN`:

```{r}
sum(varmf$VarName == "SEQN")
```

The other ones are typically variables that are recorded in multiple tables; for example,

```{r}
subset(varmf, VarName == "PHAFSTMN", select = c(Table, TableDesc, BeginYear, EndYear)) |> kable()
```

It is not immediately obvious whether these are these measuring the
same thing. If we include the variable descriptions in the above
table, we see that they are sometimes different.

```{r}
subset(varmf, VarName == "PHAFSTMN", select = c(Table, TableDesc, VarDesc)) |>
    sort(by = ~ Table) |> kable()
```


# Aggregation strategy

To make a searchable table that is more concise that the full table
but retains potentially different interpretations of the same variable
name, we will create a table with one row for each distinct
combination of `VarName` and `VarDesc`, and alongside them, list all
tables where this combination appears.


```{r}

summarizeTable <- function(d) {
    with(d, sprintf("[%s]<br>(%s)",
                    paste(sort(Table), collapse = ", "),
                    paste(sort(unique(TableDesc)), collapse = "<br> ")))
}
aggTable <- unique(varmf[c("VarName", "VarDesc", "Component")]) |>
    sort(by = ~ Component + VarName + VarDesc)
aggTable$Details <- ""
aggTable <- subset(aggTable, VarName != "SEQN") # not useful
for (i in seq_len(nrow(aggTable))) {
    if (interactive() && i %% 100 == 0) cat("\r", i, " / ", nrow(aggTable), "      ")
    aggTable$Details[[i]] <-
        summarizeTable(subset(varmf,
                              VarName == aggTable$VarName[[i]] &
                              VarDesc == aggTable$VarDesc[[i]]))
}
```

We can now either save this is a CSV file for further processing, e.g.,

```r
write.csv(aggTable, file = "nhanes-variables.csv", row.names = FALSE)
```

or use the `DT` package to create a searchable HTML table using the
DataTables Javascript library.

```{r}
library(DT)
dt <- datatable(aggTable,
                rownames = FALSE,
                colnames = c("Variable", "Description", "Component", "Source"),
                escape = FALSE, editable = FALSE,
                options = list(columnDefs = list(
                                   list("searchable" = FALSE,
                                        "targets" = c(2, 3))
                               )))
saveWidget(dt, file = "nhanes-variables.html", selfcontained = TRUE)
```

Both of these will create fairly large files, so they are not actually
created here. For demonstration the following gives the table only for
the Demographics component, which is the smallest.

```{r}
demo <-
    DT::datatable(subset(aggTable, Component == "Demographics"),
                  rownames = FALSE,
                  colnames = c("Variable", "Description", "Component", "Source"),
                  escape = FALSE, editable = FALSE,
                  options = list(columnDefs = list(
                                     list("searchable" = FALSE,
                                          "targets" = c(2, 3))
                                 )))
saveWidget(demo, file = "nhanes-variables-demographic.html", selfcontained = TRUE)
```

The result is available [here](nhanes-variables-demographic.html).



# Enhancements

Though useful, the information available in the variable manifest is
limited. More useful information is available in the per-table
codebooks (which can be accessed using the `nhanesCodebook()`
function and summarized using the `nhanesTableSummary()` function),
including a usually shorter description of each variable referred to
as the "SAS Label", along with information about possible response
values and their counts, including the number of missing values.

Such information can be easily incorporated into the aggregate table
created above, but requires processing the codebooks for all available
tables, which involves downloading a large number of files. Even more
useful information (such as how many unique non-missing values are
actually available for a given variable) involves inspecting the
actual data files, which again involves downloading and processing a
large number of files. These steps are less time-consuming if one uses
the dockerized version of the package where both the codebooks and
datasets are available through a local database.



See <scripts/generate-variable-table.R>


